---
title: "DESeq2 Data Analysis"
author: "Rayna M Harris"
date: "January 23, 2017"
output:
  md_document:
    variant: markdown_github
---

## RNAseq gene expression analysis with DESeq2 

This workflow was modified from the DESeq2 tutorial found at: https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.pdf

First I load a handful of packages for data wrangling, gene expression analysis, data visualization, and statistics.
```{r setup, message=F, warning=F}
library(dplyr) ## for filtering and selecting rows
library(plyr) ## for renmaing factors
library(reshape2) ## for melting dataframe
library(DESeq2) ## for gene expression analysis
library(edgeR)  ## for basic read counts status
library(magrittr) ## to use the weird pipe
library(genefilter)  ## for PCA fuction

## Functions
source("functions_RNAseq.R")
source("resvalsfunction.R")

## set output file for figures 
knitr::opts_chunk$set(fig.path = '../figures/02_RNAseq/')

```


Now, I create data frames from three csv files
- count: Contains counts for all transcripts generated from the program Kallisto. This data can be reproducibed from the file kallisto.Rmd
- geneids: Contains the ensemble ids and gene names for all the transcripts in the counts data frame. This file will be used to convert transcipt counts to gene counts. This file was also created via kallisto.Rmd file
- Traits: This file contains all the information I collected for each sample that was sequenced. Not all columns will be needed, so some are removed later. 

```{r getdata, message=F}
count <- read.csv("../data/02_count.csv", row.names=1, check.names=FALSE )
geneids <- read.csv("../data/02_geneids.csv")
Traits <- read.csv("../data/02_JA16444samples.csv", sep=",", header = TRUE, stringsAsFactors=FALSE, na.string = "NA")
```

Rather than analyze transcript level counts, I want to examine gene-level counts. In this next section, I do some data wrangling to tidy the data and summarize the counts for all transcripts belonging to the same gene. 

```{r countbygene}
countbygene <- full_join(geneids, count) # merge count and gene id dataframes
countbygene <- countbygene[-c(1:6,8:12)] # remove unnecesary columns (aka, keep gene name and counts for samples)
countbygene <- melt(countbygene, id=c("gene")) # lenghten 
countbygene$variable <- gsub('.{4}$', '', countbygene$variable) # string split to remove last for characters aka "_S##"
countbygene$variable <- gsub("\\_", "-", countbygene$variable) #replace _ with - in name
countbygene  <- dcast(countbygene, gene ~ variable, value.var= "value", fun.aggregate=mean) # widen by sum
row.names(countbygene) <- countbygene$gene # make gene the row name 
countbygene[1] <- NULL # now remove the gene name from the df
countbygene <- round(countbygene) # round all value to nearest 1s place
head(countbygene)
summary(countbygene)
```

In this next section, I tidy the trait data for each sample so that I can calculate differential gene expression for the traits of interest. I also remove some samples for reasons described within the code blocks.

```{r traits}
rownames(Traits) <- Traits$RNAseqID    # set $genoAPAsessionInd as rownames
Traits <- Traits[c(1,3,5:6,10:11)]  #keeping informative volumns
Traits <- Traits %>% dplyr::filter(!grepl("100|101|147-|148-|147D-CA1-1|145B-CA3-1|146C-CA3-4", RNAseqID))  
# remove 100, 100, 147, and 148:  homecage animals
# Remove 147D_CA1_1 and 145B_CA3_1: bad samples with no reads.
# Remove 146C-CA3-4: outlier on all pc analyses
Traits$APAconflict <- as.factor(paste(Traits$APA, Traits$Conflict, sep="_")) # adding combinatorial traits columns
Traits$ID <- gsub("[[:punct:]]", "", Traits$Mouse) #make a column that thas id without the dash
row.names(Traits) <- Traits$RNAseqID # make gene the row name 
Traits$APA <- NULL ## delete old APA column
names(Traits)[names(Traits)=="APAconflict"] <- "APA" #rename APAconflict APA (for simplicity)

# rename factors & group all Control animals into 1 group
Traits$APA <- revalue(Traits$APA, c("Trained_Conflict" = "Conflict")) 
Traits$APA <- revalue(Traits$APA, c("Trained_NoConflict" = "Consistent")) 
Traits$APA <- revalue(Traits$APA, c("Yoked_Conflict" = "Control")) 
Traits$APA <- revalue(Traits$APA, c("Yoked_NoConflict" = "Control")) 
head(Traits)
```

Now, we are ready to calculate differential gene expression using the DESeq package. For simplicity, I will use the standard nameing of "countData" and "colData" for the gene counts and gene information, respectively.

```{r prepforDESeq2}
countData <- countbygene #set the countdata to be the countbygene df
colData <- Traits %>% arrange(RNAseqID) #set the coldata to be the countbygene df

## colData and countData must contain the exact same sample. I'll use the next three lines to make that happen
savecols <- as.character(colData$RNAseqID) #select the sample name column that corresponds to row names
savecols <- as.vector(savecols) # make it a vector
countData <- countData %>% dplyr::select(one_of(savecols)) # select just the columns that match the samples in colData


# colData must be factors
cols = c(1:4,7)
colData[,cols] %<>% lapply(function(x) as.factor(as.character(x)))
colData$Slice <- as.factor(colData$Slice)
str(colData)
```

Now, we are create differential gene expression object and remove genes with 0 counts. Before filtering, there are 22,485 genes in the object. After filtering genes with 0 counts, we will be left with 17,746 genes that are expressed in a least 1 sample. Then, we can caluate the size factors, estimate gene dispersion estimates, fit a model, test for outliers, and remove outliers.

```{r DESeq2}
## create DESeq object using the factors Punch and APA
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ Punch + APA + Punch*APA)

## DESeq2 1.3.7 specify the factor levels
dds$Punch <- factor(dds$Punch, levels=c("DG","CA3", "CA1"))
dds$APA <- factor(dds$APA, levels=c("Control", "Consistent", "Conflict"))


dds # view the DESeq object - note numnber of genes

## DESeq2 1.3.6 Pre-filtering genes with 0 counts
dds <- dds[ rowSums(counts(dds)) > 1, ] 

dds # view the DESeq object - note numnber of genes
# dim: 17746 45 
# 17,746 genes and 45 samples

## DESeq2 1.4  Differential expression analysi
dds <- DESeq(dds)

## for variance stablized gene expression and log transformed data
rld <- rlog(dds, blind=FALSE)
#vsd <- varianceStabilizingTransformation(dds, blind=FALSE)
#vsd.fast <- vst(dds, blind=FALSE)
```

Now, we can calculate the number of differentiall expressed genes for each contrast. See the function in "resvalsfunction.R" for details.

```{r DESeq2contrasts}
source("resvalsfunction.R")
contrast1 <- resvals(contrastvector = c("Punch", "CA1", "DG"), mypadj = 0.05) #2497
contrast2 <- resvals(contrastvector = c("Punch", "CA1", "CA3"), mypadj = 0.05) #1803
contrast3 <- resvals(contrastvector = c("Punch", "CA3", "DG"), mypadj = 0.05) #3445
contrast4 <- resvals(contrastvector = c("APA", "Consistent", "Control"), mypadj = 0.05) #95
contrast5 <- resvals(contrastvector = c("APA", "Conflict", "Control"), mypadj = 0.05) #42
contrast6 <- resvals(contrastvector = c("APA", "Conflict", "Consistent"), mypadj = 0.05) # 0 

#create a new DF with the gene counts
## note: contrast1 had 0 differentially expressed genes, so it is not included 
rldpadjs <- assay(rld)
rldpadjs <- cbind(rldpadjs, contrast1, contrast2, contrast3, contrast4, contrast5, contrast6)
rldpadjs <- as.data.frame(rldpadjs)
rldpadjs <- rldpadjs[ , grepl( "padj" , names( rldpadjs ) ) ]
head(rldpadjs)
```

```{r dataforvolcanoplots}
volcano1 <- respadjfold(contrastvector = c("Punch", "CA1", "DG")) 
volcano2 <- respadjfold(contrastvector = c("Punch", "CA1", "CA3")) 
volcano3 <- respadjfold(contrastvector = c("Punch", "CA3", "DG")) 
volcano4 <- respadjfold(contrastvector = c("APA", "Consistent", "Control")) 
volcano5 <- respadjfold(contrastvector = c("APA", "Conflict", "Control")) 
volcano6 <- respadjfold(contrastvector = c("APA", "Conflict", "Consistent")) 

volcanos <- cbind(volcano1,volcano2, volcano3, volcano4, volcano5, volcano6)
volcanos <- as.data.frame(volcanos)
```

Now let's look at a heat map of the data

```{r heatmap}
DEGes <- assay(rld)
DEGes <- cbind(DEGes, contrast1, contrast2, contrast3, contrast4, contrast5, contrast6)
DEGes <- as.data.frame(DEGes) # convert matrix to dataframe
DEGes$rownames <- rownames(DEGes)  # add the rownames to the dataframe
DEGes$padjmin <- with(DEGes, pmin(padjPunchCA1DG, padjPunchCA1CA3, padjPunchCA3DG, padjAPAConsistentControl, padjAPAConflictControl, padjAPAConflictConsistent)) # create new col with min padj
DEGes <- DEGes %>% filter(padjmin < 0.05)
rownames(DEGes) <- DEGes$rownames
drop.cols <-colnames(DEGes[,grep("padj|pval|rownames", colnames(DEGes))])
DEGes <- DEGes %>% dplyr::select(-one_of(drop.cols))
DEGes <- as.matrix(DEGes)
DEGes <- DEGes - rowMeans(DEGes)
head(DEGes)

## the heatmap annotation file
df <- as.data.frame(colData(dds)[,c("Punch","APA")]) ## matrix to df

```

Now lets look at a principle component analysis of the data

```{r pca}
# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(rld, intgroup=c("Punch","APA"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
percentVar

## statistics
aov1 <- aov(PC1 ~ Punch, data=pcadata)
summary(aov1) 
TukeyHSD(aov1, which = "Punch") 

aov2 <- aov(PC2 ~ Punch, data=pcadata)
summary(aov2) 
TukeyHSD(aov2, which = "Punch") 

aov3 <- aov(PC3 ~ APA, data=pcadata)
summary(aov3) 
TukeyHSD(aov3, which = "APA")

aov4 <- aov(PC4 ~ APA, data=pcadata)
summary(aov4) 
TukeyHSD(aov4, which = "APA") 

lm4 <- lm(PC4~APA*Punch, data=pcadata)
summary(lm4)
anova(lm4) 

lm124 <- lm(PC1+PC2+PC4~APA*Punch, data=pcadata)
summary(lm124)
anova(lm124) 
```

```{r writecsv}
write.csv(colData, file = "../data/02a_colData.csv", row.names = F)
write.csv(countData, file = "../data/02a_countData.csv", row.names = T)
write.csv(rldpadjs, file = "../data/02a_rldpadjs.csv", row.names = T)
write.csv(DEGes, file = "../data/02a_DEGes.csv", row.names = T)
write.csv(df, file = "../data/02a_df.csv", row.names = F)
write.csv(pcadata, file = "../data/02a_pcadata.csv", row.names = F)
write.table(percentVar, file = "../data/02a_percentVar.txt")
```

## make volcano plos here.. still perfecting
```{r volcanoplots}

res <- results(dds, contrast =c("Punch", "CA1", "DG"), independentFiltering = F)
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="DG - CA1", xlim=c(-8,8)))
with(subset(res, log2FoldChange>0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#7570b3")))
with(subset(res, log2FoldChange<0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#d95f02")))
with(subset(res, padj>.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="grey"))
#abline(h=(-log10(0.05)),  col = "gray60")

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("Punch"))

res <- results(dds, contrast =c("Punch", "CA1", "CA3"), independentFiltering = F)
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="CA3-CA1", xlim=c(-8,8)))
with(subset(res, log2FoldChange>0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#7570b3")))
with(subset(res, log2FoldChange<0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#1b9e77")))
with(subset(res, padj>.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="grey"))
#abline(h=(-log10(0.05)),  col = "gray60")

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("Punch"))


res <- results(dds, contrast =c("Punch", "CA3", "DG"), independentFiltering = F)
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="DG - CA3", xlim=c(-8,8)))
with(subset(res, log2FoldChange>0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#1b9e77")))
with(subset(res, log2FoldChange<0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#d95f02")))
with(subset(res, padj>.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="grey"))
#abline(h=(-log10(0.05)),  col = "gray60")

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("Punch"))


res <- results(dds, contrast =c("APA", "Conflict", "Consistent"), independentFiltering = F)
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Consistent - Conflict", xlim=c(-8,8), ylim=c(0,14)))
with(subset(res, log2FoldChange>0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#ca0020")))
with(subset(res, log2FoldChange<0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#f4a582")))
with(subset(res, padj>.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="grey"))
#abline(h=(-log10(0.05)),  col = "gray60")

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("APA"))


res <- results(dds, contrast =c("APA", "Conflict", "Control"), independentFiltering = F)
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Control - Conflict", xlim=c(-8,8), ylim=c(0,14)))
with(subset(res, log2FoldChange>0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#ca0020")))
with(subset(res, log2FoldChange<0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#404040")))
with(subset(res, padj>.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="grey"))
#abline(h=(-log10(0.05)),  col = "gray60")

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("APA"))

res <- results(dds, contrast =c("APA", "Consistent", "Control"), independentFiltering = F)
with(res, plot(log2FoldChange, -log10(pvalue), pch=20, main="Control - Consistent", xlim=c(-8,8), ylim=c(0,14)))
with(subset(res, log2FoldChange>0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#f4a582")))
with(subset(res, log2FoldChange<0), points(log2FoldChange, -log10(pvalue), pch=20, col=c("#404040")))
with(subset(res, padj>.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="grey"))
#abline(h=(-log10(0.05)),  col = "gray60")

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("APA"), transform = T)
```
