---
title: "02g_RNAseqDG"
author: "Rayna M Harris"
date: "8/22/17"
output: md_document
---

```{r setup, message=F}
library(ggplot2) ## for awesome plots!
library(cowplot) ## for some easy to use themes
library(dplyr) ## for filtering and selecting rows
library(car) ## stats
library(VennDiagram) ## venn diagrams
library(pheatmap) ## awesome heatmaps
library(viridis) # for awesome color pallette
library(reshape2) ## for melting dataframe
library(DESeq2) ## for gene expression analysis
library(edgeR)  ## for basic read counts status
library(magrittr) ## to use the weird pipe
library(genefilter)  ## for PCA fuction

## load functions 
source("figureoptions.R")
source("functions_RNAseq.R")

## set output file for figures 
knitr::opts_chunk$set(fig.path = '../figures/02g_RNAseqDG/')
```

## Experimental Design

Given the large variance due to hippocampal subfield, I will subset the data by subfield to look for variation between behavioral treatment groups. This script focuses on DG.

```{r data, message=F, warning=F, echo=FALSE}
colData <- read.csv("../data/02a_colData.csv", header = T)
countData <- read.csv("../data/02a_countData.csv", header = T, check.names = F, row.names = 1)

colData <- colData %>% 
  filter(Punch %in% c("DG"))  %>% 
  droplevels()

savecols <- as.character(colData$RNAseqID) 
savecols <- as.vector(savecols) 
countData <- countData %>% dplyr::select(one_of(savecols)) 

colData %>% select(APA2,Punch)  %>%  summary()
```

## Overall differential Gene Expression analysis

```{r DESeq2, echo = F}
## create DESeq object using the factors Punch and APA
dds <- DESeqDataSetFromMatrix(countData = countData,
                              colData = colData,
                              design = ~ APA2)

dds # view the DESeq object - note numnber of genes
dds <- dds[ rowSums(counts(dds)) > 1, ]  # Pre-filtering genes with 0 counts
dds # view number of genes afternormalization and the number of samples
dds <- DESeq(dds) # Differential expression analysis
rld <- rlog(dds, blind=FALSE) ## log transformed data
vsd <- varianceStabilizingTransformation(dds, blind=F) ## variance stabilized
colnames(vsd) = colData$RNAseqID 
head(vsd)
```

## Principle component analysis

```{r pca, echo = F, message=F, warning=F}
# create the dataframe using my function pcadataframe
pcadata <- pcadataframe(rld, intgroup=c("Punch","APA2"), returnData=TRUE)
percentVar <- round(100 * attr(pcadata, "percentVar"))
percentVar

aov1 <- aov(PC1 ~ APA2, data=pcadata)
summary(aov1) 
TukeyHSD(aov1, which = "APA2") 

aov2 <- aov(PC2 ~ APA2, data=pcadata)
summary(aov2) 
TukeyHSD(aov2, which = "APA2") 

aov3 <- aov(PC3 ~ APA2, data=pcadata)
summary(aov3) 

aov4 <- aov(PC4 ~ APA2, data=pcadata)
summary(aov4) 
```

# Number of differentially expressed genes per two-way contrast

```{r Twowaycontrasts3}
#calculate significance of all two way comparisions
# see source "functions_RNAseq.R" 
contrast1 <- resvals(contrastvector = c("APA2", "consistent", "yoked_consistent"), mypval = 0.05) # 74
contrast2 <- resvals(contrastvector = c("APA2", "conflict", "yoked_conflict"), mypval = 0.05) # 2
contrast3 <- resvals(contrastvector = c("APA2", "conflict", "consistent"), mypval = 0.05) # 0
contrast4 <- resvals(contrastvector = c("APA2", "yoked_conflict", "yoked_consistent"), mypval = 0.05) # 2
```

## venn diagrams

```{r venndiagrams, echo=F}
rldpadjs <- assay(rld)
rldpadjs <- cbind(rldpadjs, contrast1, contrast2, contrast4)
rldpadjs <- as.data.frame(rldpadjs)
rldpadjs <- rldpadjs[ , grepl( "padj" , names( rldpadjs ) ) ]

venn1 <- row.names(rldpadjs[rldpadjs[1] <0.05 & !is.na(rldpadjs[1]),]) 
venn2 <- row.names(rldpadjs[rldpadjs[2] <0.05 & !is.na(rldpadjs[2]),]) 
venn3 <- row.names(rldpadjs[rldpadjs[3] <0.05 & !is.na(rldpadjs[3]),]) 

candidates <- list("consistent vs yoked" = venn1 , "conflict vs yoked" = venn2,  "yoked vs yoked" = venn3)

prettyvenn <- venn.diagram(
  scaled=T,
  x = candidates, filename=NULL, 
  col = "black",
  fill = c( "white", "white", "white"),
  alpha = 0.5,
  cex = 1, fontfamily = "sans", #fontface = "bold",
  cat.default.pos = "text",
  cat.dist = c(0.03, 0.03, 0.02), cat.pos = 1,
  cat.cex = 1, cat.fontfamily = "sans")
grid.draw(prettyvenn)
```

## Volcano plots

```{r volcanos, message=F, warning=F, echo=F}
res <- results(dds, contrast =c("APA2", "consistent", "yoked_consistent"), independentFiltering = T, alpha = 0.05)
summary(res)

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("APA2"))

resOrdered <- res[order(res$padj),]
head(resOrdered, 10)

data <- data.frame(gene = row.names(res),
                   pvalue = -log10(res$padj), 
                   lfc = res$log2FoldChange)
data <- na.omit(data)
data <- data %>%
  mutate(color = ifelse(data$lfc > 0 & data$pvalue > 1.3, 
                        yes = "consistent", 
                        no = ifelse(data$lfc < 0 & data$pvalue > 1.3, 
                                    yes = "yoked_consistent", 
                                    no = "none")))
top_labelled <- top_n(data, n = 5, wt = lfc)
# Color corresponds to fold change directionality
colored <- ggplot(data, aes(x = lfc, y = pvalue)) + 
  geom_point(aes(color = factor(color)), size = 1, alpha = 0.5, na.rm = T) + # add gene points
  theme_bw(base_size = 8) + # clean up theme
  theme(legend.position = "none") + # remove legend 
  scale_color_manual(values = volcano1) + 
  theme(panel.grid.minor=element_blank(),
           panel.grid.major=element_blank()) + 
  scale_x_continuous(name="log2 (consistent/yoked)",
                     limits=c(-4, 4)) +
  scale_y_log10(limits=c(0.00001, 100)) +
  labs( y = NULL) +
  geom_hline(yintercept = 1.3,  size = 0.25, linetype = 2 )
colored

pdf(file="../figures/02g_RNAseqDG/consistent.pdf", width=1.5, height=1.75)
plot(colored)
dev.off()

###############

res <- results(dds, contrast =c("APA2",  "conflict", "consistent"), independentFiltering = T, alpha = 0.05)
resOrdered <- res[order(res$padj),]
head(resOrdered, 10)

summary(res)

topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("APA2"))

data <- data.frame(gene = row.names(res),
                   pvalue = -log10(res$padj), 
                   lfc = res$log2FoldChange)
data <- na.omit(data)
data <- data %>%
  mutate(color = ifelse(data$lfc > 0 & data$pvalue > 1.3, 
                        yes = "conflict", 
                        no = ifelse(data$lfc < 0 & data$pvalue > 1.3, 
                                    yes = "consistent", 
                                    no = "none")))
top_labelled <- top_n(data, n = 5, wt = lfc)
# Color corresponds to fold change directionality
colored <- ggplot(data, aes(x = lfc, y = pvalue)) + 
  geom_point(aes(color = factor(color)), size = 1, alpha = 0.5, na.rm = T) + # add gene points
  theme_bw(base_size = 8) + # clean up theme
  theme(legend.position = "none") + # remove legend 
  scale_color_manual(values = volcano2) + 
  theme(panel.grid.minor=element_blank(),
           panel.grid.major=element_blank()) + 
  scale_x_continuous(name="log2 (conflict/consistent)",
                     limits=c(-4, 4)) +
  scale_y_log10(limits=c(0.00001, 100)) +
  labs( y = NULL) +
  geom_hline(yintercept = 1.3,  size = 0.25, linetype = 2 ) 
colored

#cvd_grid(colored) # to view plot for color blind 
pdf(file="../figures/02g_RNAseqDG/confict.pdf", width=1.5, height=1.75)
plot(colored)
dev.off()

###########

res <- results(dds, contrast =c("APA2",  "yoked_conflict", "yoked_consistent"), independentFiltering = T, alpha = 0.05)
resOrdered <- res[order(res$padj),]
head(resOrdered, 10)

summary(res)
topGene <- rownames(res)[which.min(res$padj)]
plotCounts(dds, gene = topGene, intgroup=c("APA2"))


data <- data.frame(gene = row.names(res),
                   pvalue = -log10(res$padj), 
                   lfc = res$log2FoldChange)
data <- na.omit(data)
data <- data %>%
  mutate(color = ifelse(data$lfc > 0 & data$pvalue > 1.3, 
                        yes = "yoked_conflict", 
                        no = ifelse(data$lfc < 0 & data$pvalue > 1.3, 
                                    yes = "yoked_consistent", 
                                    no = "none")))
top_labelled <- top_n(data, n = 5, wt = lfc)
# Color corresponds to fold change directionality
colored <- ggplot(data, aes(x = lfc, y = pvalue)) + 
  geom_point(aes(color = factor(color)), size = 1, alpha = 1, na.rm = T) + # add gene points
  theme_bw(base_size = 8) + # clean up theme
  theme(legend.position = "none") + # remove legend 
  scale_color_manual(values = volcano3) + 
  theme(panel.grid.minor=element_blank(),
           panel.grid.major=element_blank()) + 
  scale_x_continuous(name="log2 (yoked to conflict/ \n yoked to consistent)",
                     limits=c(-4, 4)) +
  scale_y_log10(limits=c(0.00001, 100)) +
  labs( y = NULL) +
  geom_hline(yintercept = 1.3,  size = 0.25, linetype = 2)
colored


#cvd_grid(colored) # to view plot for color blind 
pdf(file="../figures/02g_RNAseqDG/yoked.pdf", width=1.5, height=1.75)
plot(colored)
dev.off()
```
