---
title: "GO_MWU.Rmd"
author: "Rayna M Harris"
date: August 27, 2017
output: md_document
---

## Summary from Mikhail V. Matz https://github.com/z0on/GO_MWU
GO_MWU uses continuous measure of significance (such as fold-change or -log(p-value) ) to identify GO categories that are significantly enriches with either up- or down-regulated genes. The advantage - no need to impose arbitrary significance cutoff.

If the measure is binary (0 or 1) the script will perform a typical "GO enrichment" analysis based Fisher's exact test: it will show GO categories over-represented among the genes that have 1 as their measure. On the plot, different fonts are used to indicate significance and color indicates enrichment with either up (red) or down (blue) regulated genes. No colors are shown for binary measure analysis.The tree on the plot is hierarchical clustering of GO categories based on shared genes. Categories with no branch length between them are subsets of each other. The fraction next to GO category name indicates the fracton of "good" genes in it; "good" genes being the ones exceeding the arbitrary absValue cutoff (option in gomwuPlot). For Fisher's based test, specify absValue=0.5. This value does not affect statistics and is used for plotting only.

NOTES: This program drains memory and creates some very large intermediate files, especially for the biological process catagory. First, I run the stats from the command line to make sure its working. Once I've generated the temp files, I comment out then stats portions and recreate the plots by kniting the rmd file. 

```{r setup}
library(ape)
source("gomwu.functions.R")

# set output file for figures 
knitr::opts_chunk$set(fig.path = '../../figures/02e_RNAseq_GO/')
```


## DG consistent vs. yoked

```{r DGconsistentyoked, warning=F, message=F, echo=F}
# input files
input="DGconsistentyoked.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25)  

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.01, 
	level2=0.05, 
	level3=0.001, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("grey20","red","grey20","red") 
)
```


## CA3 consistent yoked

```{r CA3consistentyoked, warning=F, message=F, echo=F}
# input files
input="CA3consistentyoked.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25)  

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.01, 
	level2=0.05, 
	level3=0.001, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("grey20","red","grey20","red") 
)
```



## CA1 consistent yoked

```{r CA1consistentyoked, warning=F, message=F, echo=F}
# input files
input="CA1consistentyoked.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25)  

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.01, 
	level2=0.05, 
	level3=0.001, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("grey20","red","grey20","red") 
)
```




## DG conflict vs consistent 

```{r DGconflictconsistent, warning=F, message=F, echo=F}
# input files
input="DGconflictconsistent.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25) 

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.10, 
	level2=0.05, 
	level3=0.01, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("red","salmon","red","salmon")
)
```


## CA3 conflict consistent 

```{r CA3conflictconsistent, warning=F, message=F, echo=F}
input="CA3conflictconsistent.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
# perlPath="perl", 
# largest=0.1,  
# smallest=5,   
# clusterCutHeight=0.25)  


# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.01, 
	level2=0.005, 
	level3=0.001, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("red","salmon","red","salmon")
)
```



## CA1 consistent yoked

```{r CA1conflictconsistent, warning=F, message=F, echo=F, eval=F}
# input files
input="CA1conflictconsistent.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC


# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25) 

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.00001, 
	level2=0.000001, 
	level3=0.0000001, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("red","lightsalmon","red","lightsalmon")
)
```




## DG yoked vs. yoked

```{r DGyokedyoked, warning=F, message=F, echo=F}
# input files
input="DGyokedyoked.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25) 

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.10, 
	level2=0.05, 
	level3=0.01, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("grey20","grey60","grey20","grey60")
)
```


## CA3 yoked vs. yoked

```{r CA3yokedyoked, warning=F, message=F, echo=F}
# input files
input="CA3yokedyoked.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25) 

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.10, 
	level2=0.05, 
	level3=0.01, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("grey20","grey60","grey20","grey60")
)
```



## CA1 yoked vs. yoked

```{r CA1yokedyoked, warning=F, message=F, echo=F}
# input files
input="CA1yokedyoked.csv" 
goAnnotations="goAnnotations.tab" 
goDatabase="go.obo" 
goDivision="MF" # either MF, or BP, or CC

# Calculating stats
#gomwuStats(input, goDatabase, goAnnotations, goDivision,
#  perlPath="perl", 
#	largest=0.1,  
#	smallest=5,   
#	clusterCutHeight=0.25) 

# Data viz
gomwuPlot(input,goAnnotations,goDivision,
	absValue=-log(0.05,10),  
	level1=0.010, 
	level2=0.005, 
	level3=0.001, 
	txtsize=2,    
	treeHeight=0.5, 
  colors=c("grey20","grey60","grey20","grey60")
)
```