---
title: "00_kallistoCounts"
author: "Rayna M Harris"
date: "December 20, 2018"
output: md_document
---
  
```{r setup, warning=FALSE, message=FALSE}
library("dplyr") ## for filtering and selecting rows
library("plyr")  ## for renmaing factors
library("reshape2") ##  for melting dataframe
```

## Disclaimer

Hello! If you are viewing this page then hopefully you want to access some really large data files containing raw RNA transcript counts and estimates of transcripts per million. 

To obtain the data analyzed in this markdown file, I [ran kallisto](https://github.com/raynamharris/IntegrativeProjectWT2015/blob/master/UNIXworkflow/04_kallisto.md) on the Stampede Cluster at the Texas Advanced Computing Facility. The data are exported as abunance files in a subdirectory for every sample. 

These files and this analysis will take up considerable space and time. If you want to run the analysis, first download `GSE100225_IntegrativeWT2015` from this GitHub repository: https://github.com/raynamharris/MouseHippocampusRNAseqData and save them in this repo in `../data/. 

## Kallisto Gather

The kallisto output gives you read counts for sample in an abundance file for every single sample. This portion of the code goes through and finds each samples' abundance.tsv file, extracts the data, and combines it all into a dataframe. The `counts` file is unnormalized, but the `tpm` is the data after being normalized by transcripts per million. This script was developed with assistance from Anna Batthenhouse and Dennis Whylie.

Rather than examine unique transcripts, my analyses will focus on gene-level exprrssion. I use some string splitting to take the very long transcript identifying and create a `geneids` file that has all the database identifiers for each transcript. Then, I'll save the dount data.

```{r data, warning=FALSE}
setwd("../data/GSE100225/")
## this will create lists of all the samples
kallistoDirs = dir(".")
kallistoDirs = kallistoDirs[!grepl("\\.(R|py|pl|sh|xlsx?|txt|tsv|csv|org|md|obo|png|jpg|pdf)$",
kallistoDirs, ignore.case=TRUE)]

kallistoFiles = paste0(kallistoDirs, "/abundance.tsv")
names(kallistoFiles) = kallistoDirs
if(file.exists(kallistoFiles))
  kallistoData = lapply(
  kallistoFiles,
  read.table,
  sep = "\t",
  row.names = 1,
  header = TRUE
)

## this for loop uses the reduce function to make two data frame with counts or tpm from all the samples. note, only counts are used

ids = Reduce(f=union, x=lapply(kallistoData, rownames))
if (all(sapply(kallistoData, function(x) {all(rownames(x)==ids)}))) {
  count = data.frame(
    id = ids,
    sapply(kallistoData, function(x) {x$est_counts}),
    check.names = FALSE,
    stringsAsFactors = FALSE
)
  tpm = data.frame(
    id = ids,
    sapply(kallistoData, function(x) {x$tpm}),
    check.names = FALSE,
    stringsAsFactors = FALSE
)
}

## make a dataframe with the parts of the gene id as columns
geneids <- count[c(1)] 
geneids$ENSMUST <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 1)
geneids$ENSMUSG <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 2)
geneids$OTTMUSG <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 3)
geneids$OTTMUST <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 4)
geneids$transcript <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 5)
geneids$gene <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 6)
geneids$length <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 7)
geneids$structure1 <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 8)
geneids$structure2 <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 9)
geneids$structure3 <- sapply(strsplit(as.character(geneids$id),'\\|'), "[", 10)
geneids$transcript_lenght <- as.factor(paste(geneids$transcript, geneids$length, sep="_"))


genesforgo <- geneids %>% select(gene) %>% distinct(gene) %>% arrange(gene)
head(genesforgo)


# tpm to tpmbytranscript - note: not used for subsequent analyses
tpmbytranscript <-  full_join(geneids, tpm) # merge tpm and genids
tpmbytranscript <- tpmbytranscript[-c(1:11)]   ## keep gene name and tpm for samples)
row.names(tpmbytranscript) <- tpmbytranscript$transcript_lenght ## make gene the row name
tpmbytranscript[1] <- NULL ## make gene the row name
tpmbytranscript <- round(tpmbytranscript) #round all value to nearest 1s place

# count to countbygene
countbygene <- full_join(geneids, count) # merge count and genids
countbygene <- countbygene[-c(1:6,8:12)]   ## rkeep gene name and counts for samples)
countbygene <- melt(countbygene, id=c("gene")) ## lenghten 
countbygene  <- dcast(countbygene, gene ~ variable, value.var= "value", fun.aggregate=sum) #then widen by sum
row.names(countbygene) <- countbygene$gene ## make gene the row name
countbygene[1] <- NULL ## make gene the row name
countbygene <- round(countbygene) #round all value to nearest 1s place

setwd("../../scripts")
write.csv(geneids, "../data/00_geneids.csv", row.names=F)
write.csv(tpmbytranscript, "../data/00_tpmbytranscript.csv", row.names=T)
write.csv(countbygene, "../data/00_CountData.csv", row.names=T)
write.csv(genesforgo, "../data/00_genesforgo.csv", row.names = F)
```

## Session Info
```{r SessionInfo}
sessionInfo()
```