---
title: "Analysis of Avoidance Behavior"
output: md_document
---


```{r setup, message=F}
## load libraries 
library(tidyverse) ## for respahing data
library(cowplot) ## for some easy to use themes
library(factoextra)  ## pca with vectors
library(FactoMineR) # more pca
library(apaTables) #  for ANOVA tables

library(png) # for ading images to plots
library(grid)  # for ading images to plots

## load user-written functions 
source("functions_behavior.R")
source("figureoptions.R")

## set output file for figures 
knitr::opts_chunk$set(fig.path = '../figures/01_behavior/')
```

## Sample sizes

The 'treatment' column describes the four behavioral treatment groups.  
The 'trial' column describes the behvioral training sessions. Here I filter by a single session to calculte the number of mice. 

```{r wrangledata, message=F}
## import output from video tracker program 
behavior <- read.csv("../data/00_behaviordata.csv", header = T)

# set levels
behavior$treatment <- factor(behavior$treatment, levels = levelstreatment)
behavior$training <- factor(behavior$training, levels = levelstraining)

# sample sizes
behavior %>% 
  filter(trial == "Hab") %>%
  select(treatment)  %>%  summary()

head(behavior)
names(behavior[9:34])
```

```{r experimentaldesign, fig.height = 2.6, fig.width = 6.69}
dfshocks <- behavior %>%
  dplyr::group_by(treatment, trialNum) %>%
  dplyr::summarise(m = mean(NumShock), 
                   se = sd(NumShock)/sqrt(length(NumShock))) %>%
  dplyr::mutate(m = round(m,0)) %>%
  dplyr::mutate(measure = "NumShock")
dfshocks


numshocks <- ggplot(dfshocks,  aes(x=, trialNum, y=m, color=treatment, label = m)) + 
    geom_errorbar(aes(ymin=m-se, ymax=m+se, color=treatment), width=.1) +
    geom_line() +
    geom_point(size = 1) +
    labs(y = "\n Number of shocks") +
    scale_x_continuous(name= "trial", 
                       breaks = c(1, 2, 3, 4, 5, 6, 7, 8, 9),
                       labels = c( "P", "T1", "T2", "T3",
                                   "Rt", "T4", "T5", "T6", "Rn")) +
    scale_alpha_continuous( breaks = c(1, 2, 3)) +
    theme_ms() +
    scale_color_manual(values = treatmentcolors,
                       name  = NULL)  +
    theme(legend.position = "none",
          legend.justification=c(0,0),
          legend.text=element_text(size=5),
          strip.text = element_text(angle = 0, hjust = 0, vjust = 0),
          axis.text.y = element_blank(),
          strip.background = element_blank()) +
  facet_wrap(~treatment, nrow = 4, labeller = treatment_labeller) +
  scale_y_continuous(breaks = c(0,10,20), limits = c(0,30)) +
  geom_text(vjust= -0.5, size=2.5)
numshocks

paradigm <- png::readPNG("../figures/00_schematics/figure_1a.png")
paradigm <- ggdraw() +  draw_image(paradigm, scale = 1)

subfields <- png::readPNG("../figures/00_schematics/figure_1c.png")
subfields <- ggdraw() +  draw_image(subfields, scale = 1)


experimentaldesign <- plot_grid(paradigm, numshocks, subfields, rel_widths = c(2,1,0.8), nrow = 1,
          label_size = 8, labels = c("(a)", "(b) ", "(c)"))
experimentaldesign


pdf(file="../figures/01_behavior/experimentaldesign.pdf", width=6.69, height=2.6)
plot(experimentaldesign)
dev.off()

pdf(file="../figures/figure_1.pdf", width=6.69, height=2.6)
plot(experimentaldesign)
dev.off()
```

# Vizualizing Mean and Standard error for avoidance behaviors

To make the point and line graphs, I must create and join some data frames, then I have a function that makes four plots with specific titles, y labels and limits.

```{r behavmeanstdev}
dfa <- behavior %>%
  dplyr::group_by(treatment, trialNum) %>%
  dplyr::summarise(m = mean(NumEntrances), 
                   se = sd(NumEntrances)/sqrt(length(NumEntrances))) %>%
  dplyr::mutate(measure = "NumEntrances")

dfb <- behavior %>%
  dplyr::group_by(treatment, trialNum) %>%
  dplyr::mutate(minutes = Time1stEntr/60) %>%
  dplyr::summarise(m = mean(minutes), 
                   se = sd(minutes)/sqrt(length(minutes))) %>%
  dplyr::mutate(measure = "Time1stEntr (min)")

dfc <- behavior %>%
  dplyr::group_by(treatment, trialNum) %>%
  dplyr::summarise(m = mean(pTimeShockZone), 
                   se = sd(pTimeShockZone)/sqrt(length(pTimeShockZone))) %>%
  dplyr::mutate(measure = "pTimeShockZone")

avoidancedf <- rbind(dfa, dfb,dfc)
head(avoidancedf)

a <- meansdplots(dfa, "NumEntrances" ,  c(0,10,20,30), c(0, 35))
b <- meansdplots(dfb, "Time1stEntr (min)",  c(0,2,4,6,8), c(0, 8))
c <- meansdplots(dfc, "pTimeShockZone", c(0,.12,.25,.37), c(0, .37 ))

a
b
c

avoidancebehaviors <- plot_grid(a + theme(legend.position = c(0.05, 0.9),
                                          legend.direction = "horizontal",
                                          legend.key.size = unit(0.25, "cm")) +
                                  guides(color = guide_legend(nrow = 2)) ,
                       b + theme(legend.position = "none"), 
                       c + theme(legend.position = "none"), nrow = 1,
                       label_size = 8,
                       labels = c("(a)", "(b)", "(c)"))
avoidancebehaviors

```




### Principle component analysis 

Next, I next reduced the dimentionality of the data with a PCA anlaysis. 

```{r PCA}
pca.all <- makepcadf(behavior)

retention <- behavior %>% filter(trialNum == 9)



pca.Rn <- pca.all %>% 
  filter(trialNum == 9) %>%  
  select(ID:PC2) 
head(pca.Rn)


pca.Rn.summary <- pca.all %>% filter(trialNum == 9) %>% 
  group_by(treatment) %>% 
  dplyr::summarize(avePC1 = mean(PC1),
                   avePC2 = mean(PC2),
                   sePC1 = sd(PC1)/sqrt(length(PC1)),
                   sePC2 = sd(PC2)/sqrt(length(PC2)))


d <- ggplot(pca.all, aes(x = PC1, y = PC2, color = treatment, fill = treatment)) +

  geom_point(data = pca.all, aes(alpha = Day)) + 
  geom_point(data = pca.Rn.summary, aes(x = avePC1, y = avePC2), size = 4) +
  theme_ms() +
    scale_fill_manual(guide = 'none',values = treatmentcolors) +
  scale_color_manual(guide = 'none',values = treatmentcolors) +
  scale_alpha_continuous(breaks = c(1, 2, 3)) +
  theme(legend.position = "none") +
  labs( x = "PC1: 38.3% variance explained",
        y = "PC2: 16.7% \n variance explained",
       subtitle = " ") 
d

# get contributions
df <- behavior %>% select(TotalPath:AnnularKurtosis)
res.pca <- PCA(df,  graph = FALSE)
# Visualize eigenvalues/variances
fviz_screeplot(res.pca, addlabels = TRUE, ylim = c(0, 50))


# Contributions of variables to PC1
e <- fviz_contrib_rmh(res.pca, choice = "var", axes = 1, top = 8, 
                 ylab = "PC1 % contrib.", xlab = "estimates of memory", subtitle = " ") +
  theme_ms() + theme(axis.text.x = element_text(angle=45, hjust = 1))
# Contributions of variables to PC2
f <- fviz_contrib_rmh(res.pca, choice = "var", axes = 2, top = 8, 
                 ylab = "PC2 % contrib." , xlab = "estimates of activity", subtitle = " ") +
  theme_ms() + theme(axis.text.x = element_text(angle=45, hjust = 1))

e
f

pcaplots <- plot_grid(d,e,f, labels = c("(d)", "(e)", "(f)"),
           nrow = 1,
           label_size = 8)
pcaplots
```

```{r avoidance, fig.width=6.69,fig.height=3.5}
avoidance <- plot_grid(avoidancebehaviors, pcaplots, nrow = 2)
avoidance

pdf(file="../figures/01_behavior/avoidance.pdf", width=6.69, height=3.5)
plot(avoidance)
dev.off()

pdf(file="../figures/figure_2.pdf", width=6.69, height=3.5)
plot(avoidance)
dev.off()
```

## now all the stats

Maybe it should be done like this:
Q1: are the groups different? 1-way ANOVA of groups on Pre
Q2: are the groups different during initial training T1-T3? 2-way ANOVA of groups X trial
Q3: do the groups differ in initial recall? 1-way ANOVA of groups on Rt
Q4: Do the groups differ in subsequent training? T4-T6 2-way ANOVA of groups X trial
Q5: do the groups differ in subsequent recall? 1-way ANOVA of groups on Rn

```{r stats}
# Q1. Are groups different at pre? No.
Q1 <- onewayANOVAfor3measures(behavior, "Hab", "Pre-training (Pre)")
  
  
# Q2. Are the groups different during initial training T1-T3? Yes (sometimes alone, sometime interaction)
T1T2T3 <-  behavior %>% filter(trial %in% c("T1", "T2", "T3"))
Q2 <- twowayANOVAfor3measures(T1T2T3, "Initial training (T1 - T3)")


# Q3 Do the groups differ in initial recall? Yes.
Q3 <- onewayANOVAfor3measures(behavior, "Retest", "Initial recall (Rt)")

# Q4 Do the groups differ in subsequent training? Yes
T4T5T6 <-  behavior %>% filter(trial %in% c("T4_C1", "T5_C2", "T6_C3"))
Q4 <- twowayANOVAfor3measures(T4T5T6,  "Conflict training  (T4 - T6)")


# Q5  Do the groups differ in subsequent recall? Yes
Q5 <- onewayANOVAfor3measures(behavior, "Retention", "Conflict recall (Rn)")


# more stats. didn't bother with a function

PC1all <- apa.aov.table(aov(PC1 ~ treatment , data=pca.all))
PC1all <- as.data.frame(PC1all$table_body)
totaldf <- PC1all[3, 3]
PC1all$df <- paste(PC1all$df, ", " , totaldf, sep = "")
PC1all$ANOVA <- "PC1 ~ treatment"
PC1all$trials <- "All trials"

PC2all <- apa.aov.table(aov(PC2 ~ treatment , data=pca.all))
PC2all <- as.data.frame(PC2all$table_body)
totaldf <- PC2all[3, 3]
PC2all$df <- paste(PC2all$df, ", " , totaldf, sep = "")
PC2all$ANOVA <- "PC2 ~ treatment"
PC2all$trials <- "All trails"

PC1rn <- apa.aov.table(aov(PC1 ~ treatment , data=pca.Rn))
PC1rn <- as.data.frame(PC1rn$table_body)
totaldf <- PC1rn[3, 3]
PC1rn$df <- paste(PC1rn$df, ", " , totaldf, sep = "")
PC1rn$ANOVA <- "PC1 ~ treatment"
PC1rn$trials <- "Retention (Rn)"

PC2rn <- apa.aov.table(aov(PC2 ~ treatment , data=pca.Rn))
PC2rn <- as.data.frame(PC2rn$table_body)
totaldf <- PC2rn[3, 3]
PC2rn$df <- paste(PC2rn$df, ", " , totaldf, sep = "")
PC2rn$ANOVA <- "PC2 ~ treatment"
PC2rn$trials <- "Retention"

PC.APA <- as.data.frame(rbind(PC1all, PC2all, PC1rn, PC2rn))
PC.APA <- PC.APA %>% 
    select(trials, ANOVA, Predictor, df, "F", p) %>%
    filter(!Predictor %in% c("(Intercept)", "Error"))

Q12345 <- as.data.frame(rbind(Q1,Q2,Q3,Q4,Q5, PC.APA))
Q12345
```



```{r PCcorrelations}

pca.nopre <- pca.all %>% filter(trialNum != 1)

cor.test(pca.nopre$PC1, pca.nopre$PC2, method = c("pearson"))

ggplot(pca.nopre, aes(PC1, PC2, color = treatment)) +
  geom_point()
```


## save files

```{r writefiles}
# supp table 1
write.csv(behavior, file = "../data/01_behavior.csv", row.names = FALSE)

# supp table 2
write.csv(avoidancedf, file = "../data/01_avoidance.csv", row.names = FALSE)

# table 1
write.csv(Q12345, "../data/01_APA.csv", row.names = F)

# unsure wether/how to include
write.csv(pca.all, file = "../data/01_pca.all.csv", row.names = FALSE)
write.csv(pca.Rn.summary, file = "../data/01_pca.Rn.summary.csv", row.names = FALSE)
write.csv(pca.Rn, file = "../data/01_pca.Rn.csv", row.names = FALSE)
```


```{r citations}
citation("tidyverse")
citation("cowplot")  
citation("factoextra")   
citation("FactoMineR")  
citation("apaTables")  
```